Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-04-03 article.tex
Remove performance mention before performance section
Move figures to the end
Add software and hardware (yoda)

\documentclass{article}
\usepackage[a4paper, margin=3cm, bottom=4cm]{geometry}

@@

\input{latex/figure1.tex}

@@

\input{latex/figure2.tex}

@@

\input{latex/figure3.tex}

@@

\input{latex/figure4.tex}

@@

Switching backends to \texttt{numba} can improve model runtime by a factor up to x30 or more for big or long running models,
however incurs in a compile time penalty for the first run that must be taken into account.

@@

\input{latex/figure5.tex}

@@

\input{latex/figure6.tex}

@@

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations using
COPASI (with BasiCO) with the LSODA solver,
RoadRunner (with Tellurium) with the comparable CVODE solver, and
SimBio with the LSODA solver.
For SimBio, we considered three variants:
NumPy backend and scipy's LSODA solver,
Numba backend and scipy's LSODA solver,
Numba backend and numbalsoda's LSODA solver.
In all cases, we used a relative tolerance of \(10^{-9}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

For COPASI and Tellurium,
we noted that its runtime depended on the number of evaluation points,
something that does not seem to happen with SimBio (Figure~\ref{fig-runtime}, left).
While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
we obtained an order of magnitude speed-up using the numba backend
putting it on par with them.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids the Python interpreter
between each of the integration steps.
This last combination beats all other methods,
which is also true for the other models we tested (Figure~\ref{fig-runtime}, right).

@@

>>>>

\documentclass{article}
\usepackage[a4paper, margin=3cm]{geometry}

@@

@@

@@

@@

@@

@@

@@

@@

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations
on a Intel Xeon CPU E5-2620 v4
using
COPASI v4.42.284
(with BasiCO v0.64) with the LSODA solver,
RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
SimBio v0.3.2 with the LSODA solver.
For SimBio, we considered three variants:
NumPy (v1.26.4) backend and scipy (v1.12)'s LSODA solver,
Numba (v0.59.0) backend and scipy's LSODA solver,
Numba backend and numbalsoda's LSODA solver.
In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.

@@

\newpage

\input{latex/figure1.tex}
\input{latex/figure2.tex}
\input{latex/figure3.tex}
\input{latex/figure4.tex}
\input{latex/figure5.tex}
\input{latex/figure6.tex}

### 2024-04-03 latex/figure6.tex
Use vector file for figure 6
Add software and hardware (yoda)

{\centering \includegraphics{src/performance/figures/performance.png}

@@

\caption{\label{fig-runtime}Performance of different softwares to solve
models from the curated section of BioModels. (left) Run time for the
model BIOMD3 as a function of the number of output points. (right) Run
time for different models for 300 output points, using the geometric
mean of the different softwares to order them.}

>>>>

{\centering \includegraphics{src/performance/figures/performance}

@@

\caption{
    \label{fig-runtime}
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a neglibible errorbar given by the interquartile range.
}
